---
title: "Exercises Day 2"
author: "answered by Elliot Gardner, written by Richard Paquin Morel, adapted from exercises by Christina Maimone"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: FALSE
---


```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

## Load the data

Load the `gapminder` dataset.

```{r}
gapminder <- read.csv(here::here("data","gapminder5.csv"), stringsAsFactors=FALSE)
```

```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
```


## If Statement

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

Hint: use the `any` function.

```{r}
year <- 2002
if (any(gapminder$year == year)){
  print(paste0("There is data from ", year))
} else {
  print(paste0("There are not records from ", year))
}

year <- 2012
if (any(gapminder$year == year)){
  print(paste0("There is data from ", year))
} else {
  print(paste0("There are not records from ", year))
}
```

```{r}
year<-2002
if(any(gapminder$year == year)){
   print(paste("Record(s) for the year",year,"found."))
} else {
  print(paste("No records for year",year))
}
```


## Loop and If Statements

Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

Write a script that loops through the `gapminder` data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{r}
mean_pop <- mean(gapminder$pop)
countries <- unique(gapminder$country)
for (i in countries) {
  country_mean_pop <- mean(gapminder$pop[gapminder$country == i])
  
  if (country_mean_pop < mean_pop){
    country_mean_le <- mean(gapminder$lifeExp[gapminder$country == i])
    print(paste0("Country: ",i,", Mean Life Expectancy: ",country_mean_le))
  }
}
```

```{r}
overall_mean <- mean(gapminder$pop)

for (i in unique(gapminder$country)) {
   country_mean <- mean(gapminder$pop[gapminder$country==i])
   
   if (country_mean < overall_mean) {
     mean_le <- mean(gapminder$lifeExp[gapminder$country==i])
     print(paste("Mean Life Expectancy in", i, "is", mean_le))
   } 
} # end for loop
```


```{r}
conts <- unique(gapminder$continent)

for (i in conts){
  cont_mean_le <- mean(gapminder$lifeExp[gapminder$continent == i])
  
  if (cont_mean_le < 50) {
    print(paste0(i,"'s life expectancy is less than 50"))
  } else if (cont_mean_le <= 70) {
    print(paste0(i,"'s life expectancy is between 50 and 70"))
  } else {
    print(paste0(i,"'s life expctancy is greater than 70"))
  }
}
```

```{r}
lower_threshold <- 50
upper_threshold <- 70
 
for (i in unique(gapminder$continent)){
    tmp <- mean(gapminder$lifeExp[gapminder$continent==i])
    
    if (tmp < lower_threshold){
        print(paste("Average Life Expectancy in", i, "is less than", lower_threshold))
    }
    else if (tmp > lower_threshold & tmp < upper_threshold){
        print(paste("Average Life Expectancy in", i, "is between", lower_threshold, "and", upper_threshold))
    }
    else {
        print(paste("Average Life Expectancy in", i, "is greater than", upper_threshold))
    }
   
}
```


## Exercise: Write Functions

Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset.  Hint: Use `mode()` or `class()` to get the class of the data in each column. Remember that `names()` or `colnames()` returns the name of the columns in a dataset.

```{r}
names_and_types <-
  function(df) {
    cols <- colnames(df)
    for (i in cols){
      type <- class(df[[i]])
      print(paste0(i," is of type ", type))
    }
  }

names_and_types(gapminder)
#Note: Some of these were taken or modified from https://www.r-bloggers.com/functions-exercises/
```

```{r}
data_frame_info <- function(df) {
  cols <- names(df)
  for (i in cols) {
    print(paste0(i, ": ", mode(df[, i])))
  }
}
data_frame_info(gapminder)
```

Create a function that given a vector will print the mean and the standard deviation of a **vector**, it will optionally also print the median. Hint: include an argument that takes a boolean (`TRUE`/`FALSE`) operator and then include an `if` statement.

```{r}
mean_sd_median <-
  function(vec, print_median = FALSE){
    m <- mean(vec)
    std_dev <- sd(vec)
    
    if(print_median == TRUE){
      med <- median(vec)
      print(paste0("Median: ",m,", Standard Deviation: ", std_dev, ", Median: ",med))
    } else {
      print(paste0("Median: ",m,", Standard Deviation: ", std_dev))
    }
  }

mean_sd_median(gapminder$pop, TRUE)
```

```{r}
vector_info <- function(x, include_median=FALSE) {
  print(paste("Mean:", mean(x)))
  print(paste("Standard Deviation:", sd(x)))
  if (include_median) {
    print(paste("Median:", median(x)))
  }
}

le <- gapminder$lifeExp
vector_info(le, include_median = F)
vector_info(le, include_median = T)
```

## Analyzing the relationship 

Use what you've learned so far to answer the following questions using the `gapminder` dataset. Be sure to include some visualizations!

1. What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)

```{r}
for (i in unique(gapminder$year)){
  log_gdpPercap <- log(gapminder$gdpPercap[gapminder$year == i])
  log_le <- log(gapminder$lifeExp[gapminder$year == i])
  plot(log_le ~ log_gdpPercap, main = paste0("Life Expectancy Based on GDP Per Capita in ", i), xlab = "GDP Per Capita", ylab = "Life Expectancy")
  fit <- lm(log_le ~ log_gdpPercap)
  lines(log_gdpPercap, predict(fit), col = "blue")
}
```
There was a steeper relationship between Life Expectancy and GDP Per Capita in the middle of the century, and the strength of the relationship decreased over time into the 90s, but at the turn of the century it became more strong.

2. Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.
```{r}
year <- 2007
for (i in unique(gapminder$continent)){
  log_gdpPercap <- log(gapminder$gdpPercap[gapminder$continent == i])
  log_le <- log(gapminder$lifeExp[gapminder$continent == i])
  plot(log_le ~ log_gdpPercap, main = paste0("Life Expectancy Based on GDP Per Capita in ", i), xlab = "GDP Per Capita", ylab = "Life Expectancy")
  fit <- lm(log_le ~ log_gdpPercap)
  lines(log_gdpPercap, predict(fit), col = "blue")
}
```

Life expectancy in Europe, the Americas, and Oceania is strongly related, but less so in Europe, because generally most countries are high in GDP and in Life Expectancy, and for Africa there is a weak relationship because life expectancy is generally poor, even in high GDP per Capita countries.